Material Methods

### 2020-04-17 doc/MatMeth.Rmd
added in another code chunk
more edits

#set wd
#load datafile
load(file="~./MLH1repo/data/Results_setup_4.16.20.RData")

@@

## Mouse Husbandry

@@

 All mice were housed UW-Madison Biotech and MSC facilities following RARC approved protocols.  A breeding colony of wild derived _Mus musculus_ mice sampled from Gough Island (GI)is maintained at UW Veterinary school facilities.
 Mice were fed on dry standard breeder chow. Some strains sunflower seeds, nestlets and larger cages were added to improve fertility and litter survival. Adult mice were euthanized by CO asphyxiation. Neonate and embryonic mice were euthanized by decapitation following the protocols approved by the Institutional Animal Care and Use Committee at the University of Wisconsin-Madison. Over the course of data collection some breeding colonies  mice were moved from facilities. Additionally the GI strain was kept at a separate facility. We tested for effects on mean MLH1 counts across difference mouse rooms/facilities and found no effect (Supplemental section).

## Tissue Collection and Immunohistochemistry

Spermatocyte spreads were collected and perpared as described in [@peterson2019] . The majority of mice used were between 5 and 12 weeks, (supplemental table). for DMC1 spreads we targeted juvinile male mice (X ages).

Because some strains have breeding issues, all mice some strains were kept in breeding pairs (before collecting, and aged out of the desired age range. We performed tests for effects on MLH1 counts and found no effects in the older male mice ages. (Also we tested other effects, including maternal age...)

The vast majority of oocyte data was collected from neonate mice between 5 to 48 hours old. This approach maintain breeding pairing and still result in prophase oocytes <(cite timeline of oocytes in neonates)>. . Precise staging of embryos via copulary plugs was not feasible in many of these wild derived strains due to their behavior. Embryonic samples were collected when pregnancy noted in females, embryos were staged based on (X table markers). No effect on mlh1 count due to age effects (pooled as embryo and neonate) was found (Supplemental section).

Meiocyte spreads were made following [@peters_1997].  whole testes and decapsulated overaires were incubated in 3ml or 300ul of hypotonic solution for 45min respectively.

After incubation, gonads were transferred to sucrose solution for mastication. 15ul of cell slurry was transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The next morning, slides were treated with a photoflow (Kodak, cite) wash.

@@

SYCP3 was used to mark  visualize the Axial element of the synaptonemal complex. Sicne all mouse chromosomes are telocentric/acrocentric, CREST was used to indicate the centromere <insert the concentrations>

Precursurs to COs, DSBs were quantified with the DMC1 staning antibody.

## Image Processing

@@

Images of cells were capture using a Zeiss Axioplan 2 microscope with AxioLab camera and Axio Vision software (Zeiss, Cambridge, UK). Preprocessing, including cropping, noise reduction, and histogram adjustments, was performed using Photoshop (v13.0).

For MLH1 pachytene quantification, cells with a full karyotype (19 acrocentric bivalents and XY for spermatocytes or 20 acrocentric bivalents for oocytes), distinct foci, and intact bivalents were included for quantification.
Image file names were anonymized before manual scoring of MLH1 or DMC1 foci.

## Analysis

@@

The means of DMC1 foci per cell from a single mouse from a subset of srains *musculus^PWD^*, *musculus^MSM^*, *domesticus^WSB^*, and *domesticus^G^* were tested by t-tests.

@@

The distributions of CO counts per cell was assessed for normality (supplemental figure). Mouse mean MLH1 count was used to get around the bad discreet nature of count data. (normality was confirmed with X distribution, supplemental figure).

```{r MLH1.analysis, echo=FALSE, include=FALSE}

#MLH1 + Bivalent

#-models (how several were used)
#the code of models
#the tests / analysis of models

>>>>

#set wd
#load datafile
load(file="~./MLH1repo/data/Results_setup_4.17.20.RData")

@@

### Mouse Husbandry

@@

 All mice were housed UW-Madison Biotech and MSC facilities following RARC approved protocols.  A breeding colony of wild derived _Mus musculus_ mice sampled from Gough Island (G) is maintained at UW Veterinary school facilities.

<Mice were fed on dry standard breeder chow. Some strains sunflower seeds, nestlets and larger cages were added to improve fertility and litter survival. Adult mice were euthanized by CO asphyxiation. Neonate and embryonic mice were euthanized by decapitation following the protocols approved by the Institutional Animal Care and Use Committee at the University of Wisconsin-Madison. Over the course of data collection some breeding colonies  mice were moved from facilities. Additionally the GI strain was kept at a separate facility.>

### Tissue Collection and Immunohistochemistry

Spermatocyte spreads were collected and perpared as described in [@peterson2019] . The majority of mice used were between 5 and 12 weeks for MLH1 and juvinille mice (12 to 15 days) for DMC1 spreads. The majority of oocyte tissue was collected from neonate mice between 5 to 48 hours old.  Meiocyte spreads were made following [@peters_1997] - and was similar for both gonads -- adjusted for volumes. Whole testes and decapsulated overaires were incubated in 3ml or 300ul of hypotonic solution for 45min respectively. After incubation, gonads were transferred to sucrose solution for mastication. 15ul of cell slurry was transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The next morning, slides were treated with a photoflow wash (Kodak, cite).

We tested for **effects** on mean MLH1 counts per cell; across mouse rooms, ages of mature males -- and differnces between embryos and neonates. No effects were found  (Supplemental section). (Supplemental table).

<Because some strains have breeding issues, all mice some strains were kept in breeding pairs (before collecting, and aged out of the desired age range. We performed tests for effects on MLH1 counts and found no effects in the older male mice ages. (Also we tested other effects, including maternal age...)>
<This approach maintain breeding pairing and still result in prophase oocytes <(cite timeline of oocytes in neonates)>.

@@

Meiocyte spreads were incubated with antibody protocol.

To visualize chromosomes/chromatin in the meiocytes -- we used markers for the centromere (CREST) and lateral element of of the SC (SYCP3). Double strand breaks (DSB) and COs were visualized with DMC1 and MLH1 respectively.

The master mix was made --

### Image Processing

@@

Images of cells were capture using a Zeiss Axioplan 2 microscope with AxioLab camera and Axio Vision software (Zeiss, Cambridge, UK). Preprocessing, including cropping, noise reduction, and histogram adjustments, was performed using Photoshop (v13.0). Image file names were anonymized before manual scoring of MLH1 or DMC1 foci using photoshop (cite).

### Analysis

@@

SC morphology was used to stage spermatocytes (as zygotene, leptotene, or pachtyene) (following X source). The means of **DMC1** foci per cell from a single mouse from a subset of srains *musculus^PWD^*, *musculus^MSM^*, *domesticus^WSB^*, and *domesticus^G^* were tested by t-tests.

@@

For **MLH1** pachytene quantification, cells with a full karyotype (19 acrocentric bivalents and XY for spermatocytes or 20 acrocentric bivalents for oocytes), distinct foci, and intact bivalents were included for quantification. (Quality score 1 to 5 )  based on foci brightness /staining and spreading of bivalents. The distributions of MLH1 counts per cell was assessed for normality (supplemental figure).

```{r MLH1.analysis, echo=FALSE, include=FALSE}
#MLH1 + Bivalent

#-models (how several were used)
#the code of models
#the tests / analysis of models
```
